---
title: "novaRush with Fluree v4: Full API Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{novaRush with Fluree v4: Full API Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette demonstrates every major operation against a locally running Fluree v4
instance. All examples target the `novaRush/demo` ledger.

**Prerequisites**: Fluree v4 installed and a server running (`fluree serve` from a
directory initialised with `fluree init`).

## Configuration

```{r config}
library(novaRush)

conf <- list(
  host    = "localhost",
  port    = 8090,
  ledger  = "novaRush/demo"
)
```

A shared `@context` saves repeating namespace prefixes on every call.

```{r context}
ctx <- list(
  schema = "http://schema.org/",
  ex     = "http://example.org/"
)
```

## Connect

```{r connect}
db <- FlureeInstance$new(conf)$setContext(ctx)$connect()
```

## Insert — add new triples

Insert is the right operation when subjects are new. Existing subjects are not
touched.

```{r insert}
db$insert(list(
  insert = list(
    list(`@id` = "ex:alice", `@type` = "schema:Person",
         `schema:name` = "Alice", `schema:age` = 30L),
    list(`@id` = "ex:bob",   `@type` = "schema:Person",
         `schema:name` = "Bob",   `schema:age` = 25L)
  )
))$send()
```

## Upsert — replace predicate values idempotently

Upsert replaces every predicate you supply on the matching subject; predicates you
omit are left unchanged. Running the same upsert twice produces the same ledger state.

```{r upsert}
db$upsert(list(
  upsert = list(
    list(`@id` = "ex:bob", `schema:age` = 26L)
  )
))$send()
```

## Update — conditional where/delete/insert

Use update when you need to bind the current value before replacing it. The `where`
clause matches existing triples; `delete` retracts them; `insert` adds the new ones.

```{r update}
db$update(list(
  where  = list(list(`@id` = "ex:alice", `schema:age` = "?age")),
  delete = list(list(`@id` = "ex:alice", `schema:age` = "?age")),
  insert = list(list(`@id` = "ex:alice", `schema:age` = 31L))
))$send()
```

## Delete — retract all triples for a subject

Delete routes through update with a wildcard pattern that retracts every triple for
the given IRI(s).

```{r delete}
db$insert(list(
  insert = list(
    list(`@id` = "ex:temp", `schema:name` = "Temp")
  )
))$send()

db$delete("ex:temp")$send()
```

## Query — JSON-LD query

Fluree's native query language uses `select`, `where`, and optional `filter`,
`groupBy`, `orderBy`, `limit`, and `reasoning` fields.

```{r query-basic}
result <- db$query(list(
  select = list("?name", "?age"),
  where  = list(list(`@id` = "?s", `schema:name` = "?name", `schema:age` = "?age")),
  orderBy = "?name"
))$send()

result
```

### With RDFS reasoning

Declare `@type` relationships and query via `rdfs:subClassOf` inference:

```{r query-reasoning}
# Load a minimal ontology first
db$insert(list(
  insert = list(
    list(`@id` = "schema:Employee", `rdfs:subClassOf` = list(`@id` = "schema:Person"))
  )
))$send()

db$insert(list(
  insert = list(
    list(`@id` = "ex:dave", `@type` = "schema:Employee", `schema:name` = "Dave")
  )
))$send()

# Query for all Persons — includes Employees via subClassOf
result <- db$query(list(
  `@context` = list(rdfs = "http://www.w3.org/2000/01/rdf-schema#"),
  select     = list("?name"),
  where      = list(list(`@id` = "?s", `@type` = "schema:Person", `schema:name` = "?name")),
  reasoning  = "rdfs"
))$send()

result
```

## SPARQL

SPARQL 1.1 SELECT queries are sent to the same `/query` endpoint with
`Content-Type: application/sparql-query`. The ledger is taken from config.

```{r sparql}
result <- db$sparql('
  PREFIX schema: <http://schema.org/>
  PREFIX ex:     <http://example.org/>
  SELECT ?name ?age
  WHERE {
    ?s a schema:Person ;
       schema:name ?name ;
       schema:age  ?age .
  }
  ORDER BY ?name
')$send()

# SPARQL responses follow the W3C SPARQL JSON format
result$results$bindings
```

## History — commit log

`$history()` returns the commit log: each entry records the t-value, commit ID,
timestamp, and the number of triples asserted and retracted.

```{r history}
log <- db$history()$send()

# Each commit
log$commits
```

Limit to the two most recent commits:

```{r history-limit}
log <- db$history(list(limit = 2L))$send()
log$commits
```

## Time-travel query

Query the ledger at a past commit using `t` in the query body. `t` is the integer
transaction counter returned by every write operation.

```{r time-travel}
# State at t = 1 (initial insert)
result <- db$query(list(
  select = list("?name", "?age"),
  where  = list(list(`@id` = "?s", `schema:name` = "?name", `schema:age` = "?age")),
  t      = list(value = 1L)
))$send()

result
```
