---
title: "LOCAL FLUREE SETUP"
author: "Christiaan Pauw, Tshepo Sotiya, Jaco Human"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-------------------------------------------------------------------------------

# Purpose and scope of this document

This document serves as a guide explaining the contents of the "flureeSetup.R" script. 
In essence this script sets up a local instance of the fluree/ledger
docker image, but more detail about each code chunk will be given within this markdown document.

The document covers the following: 

1. Setting up a docker network and running an instance of the fluree/ledger container.
2. Creating a new ledger for capturing data.
2. Instructions on how to add an authorization key to the ledger to ensure data security.
3. Elementary examples of executing fluree ledger transactions. 

-------------------------------------------------------------------------------

## Before starting

The following resources are useful for a more comprehensive outline of how to use 
*Docker* and *Fluree* and all of the associated command line prompts (the relevant ones
will be highlighted in this document).

-------------------------------------------------------------------------------

## Required packages 

The first step would be to ensure that all of the required packages are installed 
and loaded.  See the following code block for all the required packages. 

```{r message=FALSE, warning=FALSE}
library(rjson)
library(httr)
library(jsonlite)
library(tibble)
library(plyr)
```

### Building the novaRush package

For first time users the *novaRush* package together with all its associated dependencies and helper functions needs to be built locally first before it can be used.

To do this run the following:
(the code chunk is given as a comment as this only needs to be done once)

```{r message=FALSE}
#install.packages('remotes') 
# only necessary to run this if the 'remotes' package has not yet been installed previously 

#remotes::install_github("NovaInstitute/novaRush") 
# this only needs to be done once, after which the package forms part of your local library
```

If/after the package has been built successfully make sure to execute the following:
```{r message=FALSE}
library(novaRush)
```

Now that we have ensured that all the required packages have been loaded,  we move on to setting up the 
docker environment.

-------------------------------------------------------------------------------

# Docker  

### Environment variables

The following code chunk associates an environment variable with a corresponding string.
These variables can then be used in future code (this is done solely for the sake of efficiency).

```{r message=FALSE}
Sys.setenv(fluree_link = "http://localhost:8090/fdb/") # "http://localhost:8091/fluree"
Sys.setenv(privateKey = "0c7ebd0dcbdb5796ff0175757724cffeaa948794dc0cb649b2eb525e9e70e6cb")
Sys.setenv(authId = "Tf8yFnwVxvekbvwXwsEzgwUVz4YBZMgjHEL")
```

### Docker commands 

**Create a new network if one doesn't already exist**
```{r}
system("docker network ls") # view existing containers on the current Docker host
system("docker network create fluree-network") # create a new network called "fluree-network" if it doesn't already exist
```

**Run the fluree/ledger docker image**

* `-d` : this runs the container in detached mode (in the background) so the current terminal session remains usable. This also allows the user to exit the terminal session without stopping the container.
* `--name` : this is followed by a unique name for the container (which can then be used in the future to reference this specific container in `start` or `stop` commands for example). 
* `--network` : after which we specify the network on which to run the ledger container. In this case we run the container on the network that was created in the code chunk above. 
* `-e` : set environment variables. In this case set `fdb-mode` to `dev` which runs a standalone version of Fluree, with a query engine and ledger running together. Also set the variable `fdb-open-api` to `FALSE` which allows full access on the specified port for any request.  
* `-p` : used for port mapping. In the case of the fluree/ledger image the default port is "8090", but can be changed by modifying the port before the ":".  (`8091:8090` would for example map the default port "8090" to "8091")

```{r}
system("docker run -d --name fluree-ledger --network fluree-network  -e fdb-mode=dev -e fdb-open-api=false -p 8090:8090 fluree/ledger")
```

The command given above automatically pulls the latest version of the fluree/ledger docker image from docker hub if this has not yet been done. 

The docker container is now running on the port 8090 and this is also where the API endpoint resides ("http://localhost:8090/fluree")

**More useful docker commands**

*The following command can be used to display relevant information of all the containers on the Docker host* 

```{r}
system("docker ps -a")

# Alternatively the following command can be used to only display the containers currently running
#system("docker ps")
# If no information is returned there are no containers currently in use and one needs to be started using the prompt given in the code chunk below
```

*The following commands stop and start the container respectively*
```{r}
system("docker stop fluree-ledger")
system("docker start fluree-ledger")
```

-------------------------------------------------------------------------------

# Enable authentication

**Step 1: ** Firstly set the environment variable "fdb-auth=true" to allow changes to be made

```{r}

```

**Step 2: ** Run the following commands in the CLI

```{r}

```

**Step 3: ** Install node.js and npm in the root directory of this project

```{r}

```

**Step 4: ** Install npm init -y

```{r}

```

## Generate a key pair 

The following code chunk makes use of the `generateKeyPair()` helper function within the `novaRush` package to generate a private and public key which are stored in respective variables together with a corresponding authorization ID.


```{r}
#kp <- novaRush::generateKeyPair()
#writeLines(kp, "kp.txt")
#kp <- readLines("kp.txt")
#authId <- jsonlite::fromJSON(kp)$authId
#privKey <- jsonlite::fromJSON(kp)$privKey
#pubKey <- jsonlite::fromJSON(kp)$pubKey
```


???
```{r}
#Find root id
#dfRole <- getAllEntityRecords(ledgerName = "cjp/een", entityName = "_role", signQuery = TRUE)

#create auth object locally
#transactObj <- createAuthObject(id = authId,doc = "Test 123", roles = dfRole$`_id`[[1]])

#Send to fluree
#flureeTransact(ledgerName = "cjp/een", transactObject = transactObj, signQuery = FALSE)

#dfAuth <- getAllEntityRecords(ledgerName = "cjp/een", entityName = "_auth", signQuery = FALSE) %>%
#  mutate(`_id` = as.character(`_id`))
```

-------------------------------------------------------------------------------

## Transacting in Fluree

Ledger creation, data additions, deletions and modifications are all done by means of Fluree transactions. 

### Create collections 

These are the "tables" in which data can be stored in Fluree. 

```{r}
#flureeTransact(ledgerName = "cjp/een",
#               transactObject = createCollectionObject(name  = "persons",
#                                                       doc = "Collection to for all persons",
#                                                       version = 1),
#               signQuery = FALSE)

#flureeTransact(ledgerName = "cjp/test",
#               transactObject = createCollectionObject(name  = "chat",
#                                                       doc = "Collection to for all chats",
#                                                       version = 1),
#               signQuery = FALSE)

#flureeTransact(ledgerName = "authority/test",
#               transactObject = createCollectionObject(name  = "comment",
#                                                       doc = "Collection to for all comments",
#                                                       version = 1),
#               signQuery = FALSE)

#flureeTransact(ledgerName = "authority/test",
#               transactObject = createCollectionObject(name  = "artist",
#                                                       doc = "Collection to for all artists",
#                                                       version = 1),
#               signQuery = FALSE)
```

### Create predicates 

Simply put predicates are the "columns" of the data tables.  They describe the properties that the collections have. 

In the code chunk below predicates for a person are transacted to the *authority/test* ledger. 


