---
title: "Fluree the tidy way"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fluree the tidy way}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Tibbles in, tibbles out, pipes throughout. Nothing here needs you to call a `$`
method: `fluree_connect()` hands back an opaque handle, every verb takes it as its
first argument, and `collect()` gives you a tibble.

The lower-level R6 interface (`FlureeInstance` and friends) is unchanged and still
available — see `vignette("fluree_v4_api")`. It remains the way to control the
request cycle by hand.

**Prerequisites**: Fluree v4 and a running server.

```bash
cd my-project && fluree init && fluree create novaRush/demo
fluree server start --listen-addr 127.0.0.1:8090
```

## Connect

```{r connect}
library(novaRush)

con <- fluree_connect("127.0.0.1", ledger = "novaRush/demo", port = 8090,
                      context = c(schema = "http://schema.org/",
                                  ex     = "http://example.org/"))
con
#> <fluree_connection>
#>   host   127.0.0.1:8090
#>   ledger novaRush/demo
#>   signed no
#>   context 2 prefix(es): schema, ex
```

The context is merged into every later query and transaction, so prefixes need not
be repeated. To sign messages, pass a key explicitly — the keyring is never read for
you, because unlocking it can prompt and that would hang a script.

```{r signing}
con <- fluree_connect("127.0.0.1", ledger = "novaRush/demo", port = 8090,
                      sign = TRUE, private_key = getKey())
```

## Write

`fluree_insert()` takes a triple table. The `datatype` column decides what lands in
the ledger, so numbers stay numbers rather than becoming strings.

```{r insert}
people <- tibble::tribble(
  ~subject,   ~predicate,     ~object,         ~object_type, ~datatype,
  "ex:alice", "rdf:type",     "schema:Person", "uri",        NA,
  "ex:alice", "schema:name",  "Alice",         "literal",    "string",
  "ex:alice", "schema:age",   "30",            "literal",    "integer",
  "ex:alice", "schema:knows", "ex:bob",        "uri",        NA,
  "ex:bob",   "rdf:type",     "schema:Person", "uri",        NA,
  "ex:bob",   "schema:name",  "Bob",           "literal",    "string",
  "ex:bob",   "schema:age",   "25",            "literal",    "integer")

fluree_insert(con, people)
```

Any data frame reaches the same place via semanticModelR, which derives the types
from the column classes:

```{r insert-df}
triples <- semanticModelR::pivot_longer_with_type(head(iris, 10))
fluree_insert(con, triples)
```

An ordinary data frame is refused rather than guessed at, because minting IRIs from
columns is a modelling decision, not a formatting one.

`fluree_upsert()` replaces the predicates you supply and leaves the rest alone, so
running it twice changes nothing the second time. `fluree_update()` takes a full
`where`/`delete`/`insert` body for when the old value must be bound first, and
`fluree_delete()` retracts a subject.

```{r write-others}
fluree_upsert(con, people)
fluree_delete(con, "ex:bob")
```

## Read

Build the query with `fq_*` verbs, then `collect()`.

```{r query}
fluree_query(con) |>
  fq_where(`@id` = "?s", `@type` = "schema:Person",
           `schema:name` = "?name", `schema:age` = "?age") |>
  fq_select(name, age) |>
  fq_order_by(desc(age)) |>
  fq_limit(10) |>
  collect()
#> # A tibble: 2 × 2
#>   name    age
#>   <chr> <int>
#> 1 Alice    30
#> 2 Bob      25
```

Nothing is sent until `collect()`, so a partly built query is a reusable value:

```{r reuse}
adults <- fluree_query(con) |>
  fq_where(`@id` = "?s", `schema:name` = "?name", `schema:age` = "?age") |>
  fq_select(name, age)

collect(fq_limit(adults, 5))
collect(fq_order_by(adults, desc(age)))
```

### Filters

Write filters as ordinary R expressions.

```{r filter}
fluree_query(con) |>
  fq_where(`@id` = "?s", `schema:name` = "?name", `schema:age` = "?age") |>
  fq_select(name, age) |>
  fq_filter(age > 25, startsWith(name, "A")) |>
  collect()
```

This translation earns its keep. Fluree v4.1 expects filters **inside** the `where`
clause as prefix s-expressions, and gets both alternatives wrong *silently*:

| What you write | What Fluree does |
|---|---|
| `"where": [{…}, ["filter", "(> ?age 25)"]]` | filters correctly |
| `"where": [{…}, ["filter", "?age > 25"]]` | returns **zero rows**, no error |
| top-level `"filter": [...]` | **ignored**, returns everything |

`fq_filter()` always produces the first form. Use `show_query()` to confirm:

```{r show-query}
fluree_query(con) |>
  fq_where(`@id` = "?s", `schema:age` = "?age") |>
  fq_select(age) |>
  fq_filter(age > 25) |>
  show_query()
#> {
#>   "from": "novaRush/demo",
#>   "where": [
#>     { "@id": "?s", "schema:age": "?age" },
#>     [ "filter", "(> ?age 25)" ]
#>   ],
#>   "select": [ "?age" ],
#>   "@context": { "schema": "http://schema.org/", "ex": "http://example.org/" }
#> }
```

Supported translations:

| R | Fluree |
|---|---|
| `age > 25`, `age != 30` | `(> ?age 25)`, `(!= ?age 30)` |
| `age > 25 & age < 40` | `(and (> ?age 25) (< ?age 40))` |
| `!is.na(age)` | `(bound ?age)` |
| `startsWith(name, "A")` | `(strStarts ?name "A")` |
| `grepl("^A", name)` | `(regex ?name "^A")` |
| `nchar(name) > 3` | `(> (strlen ?name) 3)` |
| `age %in% c(25, 30)` | `(in ?age [25 30])` |

Bare names become query variables. To use a value from the calling environment,
unquote it with `!!`. Unrecognised function names pass straight through, so any
Fluree function can be called, and a plain string is used verbatim:

```{r filter-escape}
cutoff <- 25
fq_filter(adults, age > !!cutoff)
fq_filter(adults, "(> ?age 25)")
```

### Whole nodes

`fq_crawl()` returns nodes rather than columns — one row per subject, one column per
predicate, with references as `@id` list-columns.

```{r crawl}
fluree_query(con) |>
  fq_where(`@id` = "?s", `@type` = "schema:Person") |>
  fq_crawl("?s") |>
  collect()
```

### Reasoning and time travel

```{r reasoning}
fluree_query(con) |>
  fq_where(`@id` = "?s", `schema:name` = "?name") |>
  fq_select(name) |>
  fq_reasoning("rdfs") |>
  collect()

# a transaction number, a commit hash, or an ISO-8601 timestamp
fluree_query(con) |>
  fq_where(`@id` = "?s", `schema:name` = "?name") |>
  fq_select(name) |>
  fq_at(1) |>
  collect()

fluree_history(con, limit = 5)
```

### SPARQL

```{r sparql}
fluree_sparql(con, "
  PREFIX schema: <http://schema.org/>
  SELECT ?name ?age
  WHERE { ?s a schema:Person ; schema:name ?name ; schema:age ?age . }
  ORDER BY ?name")
```

Values are coerced from the datatypes in the response, so `?age` comes back as an
integer. If you would rather not write SPARQL by hand, the
[glitter](https://github.com/lvaudor/glitter) package composes it with
tidyverse-style verbs and `spq_assemble()` returns the string:

```{r glitter}
library(glitter)

spq_init() |>
  spq_add("?s schema:name ?name") |>
  spq_head(10) |>
  spq_assemble() |>
  fluree_sparql(con = con)
```

## Work on the graph

`fluree_graph()` returns a [tidygraph](https://tidygraph.data-imaginist.com/) graph,
so the ledger becomes something you can traverse with dplyr verbs.

```{r graph}
library(tidygraph)
library(semanticModelR)

g <- fluree_graph(con, subject_type = "schema:Person")
g
#> # A tbl_graph: 2 nodes and 1 edges
#> # Node Data: 2 × 4 (active)
#>   name     rdf_type      `schema:age` `schema:name`
#>   <chr>    <chr>         <chr>        <chr>
#> 1 ex:alice schema:Person 30           Alice
#> 2 ex:bob   schema:Person 25           Bob
#> # Edge Data: 1 × 3
#>    from    to predicate
#>   <int> <int> <chr>
#> 1     1     2 schema:knows

rdf_summary(g)
rdf_predicates(g)

g |>
  activate(nodes) |>
  mutate(degree = centrality_degree(mode = "all")) |>
  filter(degree > 0)

rdf_neighbourhood(g, "ex:alice", order = 2)
rdf_ggraph(g)
```

Because triples remain the canonical form, a graph can go straight back:

```{r graph-write}
people_only <- g |>
  activate(nodes) |>
  filter(rdf_type == "schema:Person")

fluree_insert(con, people_only)
```

## What maps to what

| Tidy verb | Underneath |
|---|---|
| `fluree_connect()` | `FlureeInstance$new()$connect()` |
| `fluree_query() |> … |> collect()` | `$query()$send()`, response parsed to a tibble |
| `fluree_insert()` / `fluree_upsert()` | `$insert()` / `$upsert()`, with the ledger and JSON-LD assembled |
| `fluree_sparql()` | `$sparql()$send()`, W3C bindings parsed to a tibble |
| `fluree_history()` | `$history()$send()` |
| `fluree_graph()` | a crawl query plus `semanticModelR::triples_to_tbl_graph()` |
